---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 120)
```

# ertte

<!-- badges: start -->
[![R-CMD-check](https://github.com/djnavarro/ertte/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/djnavarro/ertte/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/djnavarro/ertte/graph/badge.svg)](https://app.codecov.io/gh/djnavarro/ertte)
<!-- badges: end -->

Provides tools for working with exposure-response models for time-to-event
data. It is designed with a similar interface to the 
[erglm](https://github.com/djnavarro/erglm) and [emaxnls](https://github.com/djnavarro/emaxnls)
packages, and supports the plotting interface supplied by the 
[erplots](https://github.com/djnavarro/erplots) package. It
provides two model-fitting engines, one for parametric AFT regression
models and another for semi-parametric Cox proportional-hazards regression,
behind a shared, tidy interface for survival-probability prediction,
restricted mean survival time, landmark analysis, stepwise covariate
modelling, and simulation.

## Installation

You can install the development version of ertte from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("djnavarro/ertte")
```

## Example

```{r example}
library(ertte)

mod <- ertte_aft(Surv(time, event) ~ aucss, ertte_data)
mod

ertte_predict(mod, ertte_data[1:5, ], time = c(30, 60, 90))
```
