---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ertte

<!-- badges: start -->
<!-- badges: end -->

ertte is the time-to-event member of the exposure-response (E-R) package
family, alongside [erglm](https://github.com/djnavarro/erglm) (GLM E-R
models) and [emaxnls](https://github.com/djnavarro/emaxnls) (Emax /
logistic-Emax models), all of which plug into
[erplots](https://github.com/djnavarro/erplots) for visualisation. It
wraps `survival::survreg()`-based parametric accelerated failure time
(AFT) modelling behind a tidy, consistent interface: model fitting,
AIC-based distribution selection, stepwise covariate modelling, and
simulation.

## Installation

You can install the development version of ertte from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("djnavarro/ertte")
```

## Example

```{r example}
library(ertte)

mod <- ertte_model(survival::Surv(time, event) ~ aucss, ertte_data)
mod

ertte_predict(mod, ertte_data[1:5, ], time = c(30, 60, 90))
```
