---
title: "Landmark event probabilities in ertte"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ertte)
library(survival)
```

This article covers `ertte_landmark()`, `ertte`'s simplest scalar
exposure-response (E-R) reduction of a time-to-event (TTE) endpoint. It's
the shorter sibling of the [RMST article](rmst.html), which covers a more
involved reduction (restricted mean survival time) built on the same idea
-- turning a full survival curve into a single number that can be plotted
with `erplots`' ordinary E-R grammar. If you haven't already, start with
the [overview article](overview.html) for background on `ertte_aft()`/
`ertte_coxph()`.

## Why a landmark?

A fitted TTE model describes an entire survival curve $S(t \mid x)$, but
many exposure-response questions are really about one particular point on
it: *what fraction of subjects with a given exposure profile are expected
to have had the event by some fixed, clinically meaningful time* $t^*$?
("Landmark" here refers to that fixed time, not to landmark-based dynamic
prediction from repeated measurements, a different technique with the same
name.)

Reducing to a single time point converts a TTE endpoint into an ordinary
binary response, $P(\text{event by } t^*) = 1 - S(t^* \mid x)$, which
`erplots`' existing `er_plot()`/`er_vpc()` grammars -- built for scalar
exposure-response summaries -- can visualise with no TTE-specific plotting
code at all.

## How it's computed

`ertte_landmark()` doesn't implement any new survival-curve machinery of
its own. It calls [ertte_predict()] at `time = landmark_time` and
transforms the result:

$$
P(\text{event by } t^*) = 1 - S(t^*).
$$

Because this is a decreasing, monotonic transform, the confidence interval
bounds simply swap -- the upper bound on survival becomes the lower bound
on event probability, and vice versa -- with no separate calculation
needed. Whatever validity `ertte_predict()`'s interval has for a given
engine (a Wald interval on the linear predictor for `ertte_aft()`, or
`survival::survfit()`'s own log-transform interval for `ertte_coxph()`)
carries through unchanged. This also means `ertte_landmark()` inherits all
of `ertte_predict()`'s existing edge-case handling for free -- e.g. the
all-censored-Cox-model guard, or `NA` propagation for an aliased
covariate.

One consequence of delegating this way: `landmark_time` must be a single
fixed value, unlike `ertte_predict()`'s `time` argument (which accepts a
vector). A landmark is by definition one specific point in time; if you
want event probabilities at several different landmarks, call
`ertte_landmark()` once per landmark.

## A worked example

```{r}
mod_aft <- ertte_aft(Surv(time, event) ~ aucss, ertte_data)
mod_cox <- ertte_coxph(Surv(time, event) ~ aucss, ertte_data)

nd <- ertte_data[c(1, 50, 100), ]
nd[, c("id", "aucss")]
```

```{r}
ertte_landmark(mod_aft, nd, landmark_time = 90)
```

```{r}
ertte_landmark(mod_cox, nd, landmark_time = 90)
```

As with `ertte_predict()`, the two engines' point estimates and intervals
should generally agree closely when both fit the data reasonably well (as
here), but aren't expected to match exactly, since the underlying survival
curves and confidence interval constructions differ.

## Using a landmark in a visual predictive check

Passing `landmark_time` through `simulate_args` to
`erplots::er_vpc_add_simulated()` (or directly to `er_simulate()`/
`simulate.ertte_model()`) turns each simulated replicate's raw
`sim_time`/`sim_event` into a landmark-binary outcome for comparison
against the observed data:

```{r}
sim <- simulate(mod_aft, nd, nsim = 500, seed = 3157)
sim[1:3, c("id", "sim_id", "sim_time", "sim_event")]
```

Internally, this reduces each replicate to `1` (an event on or before
`landmark_time`), `0` (known to still be event-free at `landmark_time`), or
`NA` if the replicate was censored strictly before `landmark_time` -- a
genuinely ambiguous outcome, since it's not known whether that replicate
would have had the event by $t^*$ had follow-up continued. This is the same
complete-case convention used for RMST-based VPCs (see the [RMST
article](rmst.html)'s "Using RMST in a visual predictive check" section for
the full reasoning); `erplots::er_vpc_add_simulated()`'s `mean(...,
na.rm = TRUE)` aggregation excludes these `NA`s automatically.

## Limitations

- **A landmark answers a question about one time point only.** It's
  simpler and more directly interpretable than RMST, but it discards
  everything the fitted model says about the survival curve away from
  $t^*$ -- two exposure profiles with very different survival curves can
  have the same landmark probability if the curves happen to cross near
  $t^*$.
- **The choice of $t^*$ matters, and isn't estimated from the data.**
  Different landmark times can tell different (both valid) stories about
  the same fitted model, the same caveat RMST's choice of `tau` carries.
- **A replicate censored before $t^*$ contributes no information to a
  landmark VPC and is dropped, not imputed or reweighted** -- see "Using a
  landmark in a visual predictive check" above.

## Further reading

- The [RMST article](rmst.html) covers a closely related, more involved
  scalar reduction (restricted mean survival time), including a discussion
  of why it needs a genuinely per-engine calculation, unlike
  `ertte_landmark()`.
- `?ertte_landmark` and `?ertte_predict` for the full argument/return
  documentation.
