---
title: "Survival models in ertte"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ertte)
library(survival)
```

This article is the place to start if you're new to `ertte`. It walks
through what a time-to-event (TTE) endpoint looks like, a short refresher on
the two statistical ideas `ertte`'s model engines are built on (parametric
accelerated failure time models and the Cox proportional-hazards model), and
how to fit, inspect, and predict from both engines via `ertte_aft()` and
`ertte_coxph()`. It assumes general familiarity with regression modelling,
but not with survival analysis specifically.

Three other articles build on this one: [restricted mean survival
time](rmst.html) and [landmark event probabilities](landmark.html) cover
`ertte`'s two ways of reducing a TTE endpoint to a single scalar
exposure-response summary, and [stepwise covariate modelling](scm.html)
covers `ertte`'s tools for building up a covariate model.

## Time-to-event data

A time-to-event outcome records, for each subject, how long it took for
some event of interest to happen -- and, crucially, whether it was actually
observed to happen at all. `ertte_data` (a simulated dataset shipped with
the package) has the shape this kind of data usually takes:

```{r}
ertte_data[1:6, c("id", "aucss", "time", "event")]
```

`time` is the number of days until either the event occurred or the subject
was last observed; `event` is `1` if the event was actually observed by
`time`, and `0` if the subject was instead **censored** -- still event-free
the last time they were seen, but with no way to know what happened to them
afterwards (they left the study, the study ended, etc.). Both engines in
`ertte` take a formula response built with `survival::Surv(time, event)`,
which encodes exactly this: an observed time, plus whether it represents a
true event or a censoring.

Handling censored observations correctly -- using the information that a
censored subject survived at least that long, without pretending to know
what happened after -- is the entire reason TTE models exist as a distinct
class of model, rather than just regressing `time` directly.

## A quick refresher: the survival and hazard functions

Two closely related functions describe a TTE outcome's distribution.

The **survival function** $S(t) = P(T > t)$ is the probability of still
being event-free at time $t$. It starts at $S(0) = 1$ and decreases
(weakly) as $t$ increases.

The **hazard function** $h(t)$ is the instantaneous event rate at time $t$,
conditional on having survived to $t$:

$$
h(t) = \lim_{\Delta t \to 0} \frac{P(t \le T < t + \Delta t \mid T \ge t)}{\Delta t}.
$$

The two are related by $S(t) = \exp\left(-\int_0^t h(u)\, du\right)$: the
survival function is fully determined by the hazard, and vice versa. Every
model discussed below is really a model for one of these two functions
(optionally as a function of covariates $x$) -- `ertte_aft()` models $S(t
\mid x)$ directly, and `ertte_coxph()` models $h(t \mid x)$ directly, and
each can recover the other.

## Two engines, two ways of modelling $S(t \mid x)$

`ertte` fits two structurally different kinds of TTE model. Both accept the
same kind of formula and the same kind of covariates, but make different
assumptions and answer slightly different questions.

### `ertte_aft()`: parametric accelerated failure time models

An accelerated failure time (AFT) model assumes covariates act
multiplicatively on the *time scale* -- a covariate can speed up or slow
down the passage toward the event, like a "clock" running fast or slow.
Equivalently, on the log-time scale, it's an ordinary linear model:

$$
\log(T) = \mu(x) + \sigma W,
$$

where $\mu(x)$ is the linear predictor (intercept + covariate effects),
$\sigma$ is a scale parameter, and $W$ is a random variable following a
fixed "base" distribution that doesn't depend on $x$. Different choices of
base distribution give different named AFT models, all supported by
`ertte_aft()`'s `dist` argument:

| `dist` | Base distribution of $W$ | Named model |
|---|---|---|
| `"exponential"` | Standard extreme-value, fixed $\sigma = 1$ | Exponential |
| `"weibull"` (default) | Standard extreme-value | Weibull |
| `"lognormal"` | Standard normal | Log-normal |
| `"loglogistic"` | Standard logistic | Log-logistic |

Because the whole family shares this log-location-scale structure, all four
give a fully parametric, closed-form survival function

$$
S(t \mid x) = 1 - F\!\left(\frac{\log t - \mu(x)}{\sigma}\right),
$$

where $F$ is the base distribution's CDF. `ertte_aft()` wraps
`survival::survreg()`, and adds two `ertte_aft`/`ertte_model` classes on
top so it interoperates with the rest of the package (and, if installed,
with `erplots`).

```{r}
mod_aft <- ertte_aft(Surv(time, event) ~ aucss, ertte_data)
mod_aft
```

Coefficients are on the log-time scale, so `exp(coef(...))` is usually the
more interpretable quantity: a **time ratio**, i.e. the multiplicative
change in expected survival time per unit change in the covariate.

```{r}
exp(stats::coef(mod_aft))
```

Here, `aucss`'s time ratio is just under 1: higher exposure is associated
with a (slightly) *shorter* time to event, holding nothing else fixed --
consistent with `ertte_data`'s known Weibull AFT ground truth (see
`?ertte_data`).

Choosing among the four supported distributions by AIC is automated by
`ertte_select_distribution()`:

```{r}
cmp <- ertte_select_distribution(Surv(time, event) ~ aucss, ertte_data)
cmp$comparison
```

### `ertte_coxph()`: the semi-parametric Cox proportional-hazards model

The Cox model instead assumes covariates act multiplicatively on the
*hazard scale*, and -- unlike the AFT family -- makes no assumption about
the shape of the baseline hazard over time:

$$
h(t \mid x) = h_0(t) \exp(x'\beta),
$$

where $h_0(t)$ is an unspecified **baseline hazard** (the hazard for a
subject with $x = 0$, or more precisely $x$ at the values `coxph()` centers
on) and $\beta$ are the log-hazard-ratio coefficients. This is a
*semi-parametric* model: $\beta$ is estimated by maximising a partial
likelihood that doesn't require ever specifying $h_0(t)$'s functional form,
and $h_0(t)$ itself (if needed) is estimated afterwards, nonparametrically,
from the same fit (via `survival::basehaz()`, Breslow's estimator).

The name "proportional hazards" describes its central, testable assumption:
the *ratio* of hazards between any two covariate profiles, $h(t \mid
x_1)/h(t \mid x_2) = \exp((x_1 - x_2)'\beta)$, is constant over time -- it
doesn't depend on $t$. This is a genuinely different assumption from the
AFT family's "constant multiplicative effect on the time scale", and the
two coincide only for the Weibull/exponential AFT model (the only member of
the AFT family that is *also* a proportional-hazards model).

```{r}
mod_cox <- ertte_coxph(Surv(time, event) ~ aucss, ertte_data)
mod_cox
```

Coefficients are on the log-hazard-ratio scale, so `exp(coef(...))` gives a
**hazard ratio**: the multiplicative change in the instantaneous event rate
per unit change in the covariate, at any time $t$.

```{r}
exp(stats::coef(mod_cox))
```

A hazard ratio just above 1 here says higher `aucss` is associated with a
slightly *higher* instantaneous event rate at any given time -- the same
underlying exposure effect as the AFT model above, described on a different
scale (a time ratio below 1 and a hazard ratio above 1 are two consistent
ways of saying "higher exposure, worse outcome").

The proportional-hazards assumption itself can be checked with
`survival::cox.zph()`, which tests whether a covariate's effect on the
hazard actually stays constant over time (a significant result suggests it
doesn't):

```{r}
cox.zph(mod_cox)
```

Here there's no evidence against the assumption for `aucss` -- unsurprising
given `ertte_data` was simulated from a Weibull AFT ground truth, which
(as noted above) is also a proportional-hazards model.

### Choosing between them

Both engines answer closely related exposure-response questions, and for
data that's well described by a Weibull model, their conclusions should
agree in substance (as above). Some practical differences worth knowing
about:

- `ertte_aft()` gives a fully parametric survival curve defined for all $t
  > 0$, and a `dist`-dependent closed form for effects like time ratios --
  useful when extrapolating beyond the observed follow-up range, or when a
  parametric quantity like RMST needs integrating over the whole curve (see
  the [RMST article](rmst.html)).
- `ertte_coxph()` makes no assumption about the *shape* of the baseline
  hazard over time, which is often a safer default when there's no strong
  reason to believe a particular parametric family. Its cost: no
  closed-form survival curve beyond the last observed follow-up time
  (`ertte_predict()`/`ertte_fun()` hold survival flat past that point,
  rather than extrapolating a curve), and no way to interpret effects as
  time ratios.

Both share the same `"ertte_model"` superclass, so most of the package's
downstream machinery -- `ertte_predict()`, `ertte_fun()`, `simulate()`, and
the stepwise covariate modelling functions -- work identically regardless
of which engine a model came from; see "API naming: AFT vs Cox PH" in the
package's development notes for the full dispatch scheme.

## Predicting and simulating

`ertte_predict()` returns survival probabilities (with confidence
intervals) at one or more times, for one or more covariate profiles, for
either engine:

```{r}
nd <- ertte_data[c(1, 50, 100), ]
ertte_predict(mod_aft, nd, time = c(90, 180))
ertte_predict(mod_cox, nd, time = c(90, 180))
```

The two engines' confidence intervals are constructed differently under the
hood -- a Wald interval on the AFT linear predictor vs. `survfit()`'s own
log-transform interval for the Cox model -- so they're not expected to
match exactly even when the point estimates agree closely.

`ertte_fun()` instead returns a plain R function evaluating $S(t \mid x)$
at user-specified parameters/data/times, useful for counterfactual
scenarios or plugging into simulation code:

```{r}
mod_fun <- ertte_fun(mod_aft)
mod_fun(time = 90)
```

`simulate()` draws new event times (with realistic censoring) from either
fitted model, sampling coefficients from their asymptotic sampling
distribution -- the basis for `erplots`' visual predictive checks, if
`erplots` is installed:

```{r}
simulate(mod_aft, ertte_data[1:3, ], nsim = 2, seed = 8204) |>
  dplyr::select(id, aucss, sim_id, sim_time, sim_event)
```

## Where to go next

- [Restricted mean survival time](rmst.html) and [landmark event
  probabilities](landmark.html) both reduce a TTE endpoint to a single
  exposure-response scalar, for use with `erplots`' scalar E-R plotting
  grammar.
- [Stepwise covariate modelling](scm.html) covers `ertte_scm_forward()`/
  `ertte_scm_backward()` for building up a covariate model on top of either
  engine.
- `?ertte_power` covers a power-function parameterisation for continuous
  covariates (e.g. age, weight), for either engine.

## Further reading

- Kalbfleisch, J.D. and Prentice, R.L. (2002). *The Statistical Analysis of
  Failure Time Data*, 2nd edition. Wiley. A standard reference covering
  both the AFT and Cox PH model families in depth.
- Therneau, T.M. and Grambsch, P.M. (2000). *Modeling Survival Data:
  Extending the Cox Model*. Springer. Covers `cox.zph()` and the
  proportional-hazards assumption in detail.
- `survival::survreg()`/`survival::coxph()`'s own documentation, which
  `ertte_aft()`/`ertte_coxph()` wrap directly.
