---
title: "Stepwise covariate modelling in ertte"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(ertte)
library(survival)
```

This article covers `ertte`'s stepwise covariate modelling (SCM) machinery:
`ertte_scm_forward()`/`ertte_scm_backward()`, the single-term
`ertte_add_term()`/`ertte_remove_term()` helpers they're built on, and
`ertte_scm_history()` for inspecting what happened during a search. It
assumes familiarity with `ertte_aft()`/`ertte_coxph()` -- see the [overview
article](overview.html) if you haven't already read it.

## What SCM does

Stepwise covariate modelling is a common approach (particularly in
pharmacometrics) for deciding which covariates belong in a model: starting
from a base model, candidate covariate terms are added (forward selection)
or removed (backward elimination) one at a time, keeping a term only if it
improves the model fit by more than some threshold amount.

`ertte` implements this as `ertte_scm_forward()`/`ertte_scm_backward()`,
working identically for `ertte_aft()` and `ertte_coxph()` models -- both
functions dispatch internally on the model's engine, so no separate
AFT/Cox versions are needed at the user-facing level.

## Forward selection

`ertte_scm_forward(mod, candidates, threshold)` starts from a fitted model
and repeatedly tests each not-yet-included term in `candidates`, adding
whichever single term improves the fit the most (by likelihood-ratio test
p-value) at each step, provided that improvement clears `threshold`. It
stops when no remaining candidate clears the threshold.

```{r}
mod0 <- ertte_aft(Surv(time, event) ~ aucss, ertte_data)
mod1 <- ertte_scm_forward(
  mod0,
  candidates = c("sex", "dose", "age", "weight"),
  threshold = 0.01,
  seed = 4821
)
coef(mod1)
```

Every step (including unsuccessful ones) is logged, and can be inspected
with `ertte_scm_history()`:

```{r}
ertte_scm_history(mod1)
```

Each row is one *attempt* -- one candidate term tested within one
iteration of the search. `term_p_value` is the likelihood-ratio test
p-value comparing the model with and without that term; `model_updated` is
`1` for the single attempt (if any) that was actually kept at the end of
that iteration, and `0` for the rest. An iteration with every
`model_updated == 0` means no candidate cleared the threshold that round,
which is also when the search stops.

## Backward elimination

`ertte_scm_backward(mod, candidates, threshold)` works the other
direction: starting from a model that already contains the candidate
terms, it repeatedly tests removing each one, dropping whichever single
term's removal is *least* damaging (highest likelihood-ratio p-value),
provided that p-value clears `threshold`. Because removing a term should be
a more conservative decision than adding one, `ertte_scm_backward()`'s
default `threshold` (`0.001`) is stricter than `ertte_scm_forward()`'s
(`0.01`) -- the usual convention in pharmacometric SCM workflows, sometimes
run together as "forward inclusion at p < 0.01, backward elimination at
p < 0.001".

```{r}
mod_full <- ertte_aft(Surv(time, event) ~ aucss + sex + dose + age, ertte_data)
mod_back <- ertte_scm_backward(
  mod_full,
  candidates = c("sex", "dose", "age"),
  threshold = 0.001,
  seed = 6039
)
coef(mod_back)
```

## The significance test

Every comparison in `ertte`'s SCM machinery is a likelihood-ratio
Chi-squared test, via `stats::anova()` on two nested `survreg`/`coxph`
fits. Unlike the companion `erglm` package (whose GLM-based SCM needs a
family-dependent choice of test), there's no equivalent choice to make
here: a `survreg`/`coxph` model's likelihood-ratio test doesn't vary by
distribution or engine.

If a candidate term is aliased (perfectly collinear) with a term already
in the model, `anova()` reports an `NA` p-value for the comparison. Rather
than crash the search or silently select/reject the term,
`ertte_scm_forward()`/`ertte_scm_backward()` skip that candidate for the
current step (with a warning).

## Selection criteria: p-value, AIC, or BIC

By default, both `ertte_scm_forward()` and `ertte_scm_backward()` select
terms using the likelihood-ratio p-value described above, compared against
`threshold`. The `criterion` argument lets you switch to an
information-criterion-based rule instead:

```{r}
mod_aic <- ertte_scm_forward(
  mod0,
  candidates = c("sex", "dose", "age", "weight"),
  criterion = "aic",
  seed = 4821
)
coef(mod_aic)
```

```{r}
mod_bic <- ertte_scm_backward(
  mod_full,
  candidates = c("sex", "dose", "age"),
  criterion = "bic",
  seed = 6039
)
coef(mod_bic)
```

With `criterion = "aic"` or `criterion = "bic"`, a term is added (forward)
or removed (backward) only if doing so strictly decreases the chosen
information criterion relative to the current model -- if several
candidates would each improve it, the one giving the lowest resulting IC
is kept. `threshold` has no effect in this mode and is silently ignored.

The history still records everything, via a new `criterion` column that
shows which rule was in force for each step; `term_p_value` is left `NA`
for IC-driven steps, since the likelihood-ratio test isn't computed at all
when it plays no role in selection (`model_aic`/`model_bic` are always
populated regardless of `criterion`, as they always were):

```{r}
ertte_scm_history(mod_aic)
```

## Adding or removing a single term directly

`ertte_scm_forward()`/`ertte_scm_backward()` are both built on
`ertte_add_term()`/`ertte_remove_term()`, which are also exported and
useful directly when you want to test one specific term without running a
full search:

```{r}
mod2 <- ertte_add_term(mod0, ~sex)
anova(mod0, mod2)
```

```{r}
mod3 <- ertte_remove_term(mod2, ~sex)
identical(coef(mod0), coef(mod3))
```

`term` is a one-sided formula naming exactly one term -- categorical
covariates enter as factor levels, continuous covariates enter linearly by
default. Refitting happens via an internal helper that dispatches on the
model's engine and calls the matching constructor (`ertte_aft()`/
`ertte_coxph()`) directly, rather than `stats::update()` -- `update()`
doesn't work on `ertte` model objects, because the fitted object's `$call`
refers to the constructor's own local `formula`/`data` bindings, not
anything visible in the caller's frame. This is transparent to normal
usage; it only matters if you were tempted to call `update()` on an
`ertte` model yourself.

## Power-function covariates

Continuous covariates don't have to enter a candidate set linearly.
`ertte_power(x)` (see `?ertte_power` for the full derivation) reparametrises
a continuous covariate as a power function -- `T = T_ref * (x / ref)^theta`
on the AFT time scale, or `h(t | x) = h0(t) * (x / ref)^theta` on the Cox
hazard scale -- which, after taking logs, is exactly a linear term in
`log(x / ref)`. This means it slots into `ertte_add_term()`/
`ertte_scm_forward()`/`ertte_scm_backward()` with no special handling: term
handling throughout `ertte` works generically on formula term-labels, so
`~ ertte_power(age)` (or `"ertte_power(age)"` in a `candidates` vector) is
already a legitimate term/candidate.

```{r}
mod4 <- ertte_add_term(mod0, ~ ertte_power(age))
coef(mod4)["ertte_power(age)"]
```

The fitted coefficient on `ertte_power(age)` *is* the power exponent
$\theta$ directly, and its ordinary Wald confidence interval (from
`confint()`) is already the confidence interval on $\theta$ -- no delta
method or profile likelihood needed. `ertte_power()` requires every
non-missing value of its input to be strictly positive, which is why it's
suited to covariates like age or weight rather than an exposure metric
with a placebo/zero-dose group (e.g. `aucss`, which has `0`s for the
placebo arm in `ertte_data`).

A candidate set can freely mix plain linear terms and power terms for
different (or even the same) underlying variable -- `ertte` doesn't reason
about variable semantics, so combining `age` and `ertte_power(age)` in the
same candidate set, or the same model, is left to the analyst's judgement.

## Summary of conventions

- **The default `criterion = "p-value"` uses thresholds on a
  likelihood-ratio Chi-squared test** -- this matches the classic
  pharmacometric SCM procedure. Set `criterion = "aic"` or `"bic"` to
  select on an information criterion instead, in which case `threshold`
  is ignored. `ertte_scm_history()`'s `model_aic`/`model_bic` columns are
  always logged at every step regardless of `criterion`, so you can
  inspect the search through an information-criterion lens after the fact
  even when it wasn't what drove selection.
- **`seed` only affects the order candidates are tested within a step**,
  which only matters in the essentially measure-zero case of an exact
  p-value tie between competing candidates. Model fitting itself is
  deterministic given a formula and dataset.
- **An aliased candidate (`NA` p-value) is skipped with a warning**, not
  treated as either a pass or a failure.
- **Both engines share the same SCM code path** via the internal
  `.ertte_refit()` dispatch described above -- there's nothing
  engine-specific to configure when switching between `ertte_aft()` and
  `ertte_coxph()` models.

## Further reading

- `?ertte_scm` and `?ertte_term` for the full argument/return
  documentation.
- `?ertte_power` for the derivation behind the power-function
  parameterisation.
- The [overview article](overview.html) for background on the AFT/Cox PH
  model families these functions build on.
